%% -*- coding:utf-8 -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%   $RCSfile: grammatiktheorie-include.tex,v $
%%  $Revision: 1.13 $
%%      $Date: 2010/11/16 08:40:32 $
%%     Author: Stefan Mueller (CL Uni-Bremen)
%%    Purpose: 
%%   Language: LaTeX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{localmetadata.tex}
\input{localpackages.tex}
\input{localcommands.tex}

\bibliography{bib-abbr,biblio}

\begin{document}


\frontmatter

\maketitle

\tableofcontents


\chapter*{Preface}

This book motivates an analysis of the German clause in which the verb in initial position (verb
first or verb second) is related to a trace in final position. Such analyses involving so"=called
verb movement are standard in Mainstream Generative Grammar but are frowned upon by all those
researchers that want to avoid empty elements. Working in the framework of Head-Driven Phrase
Structure Grammar I followed a linearization"=based approach \citep{Reape90a,Reape94a} from
1993--2003. In the year 2000 I noticed data that looked as if more than one constituent was fronted,
which is problematic for theories of German, since German is regarded as a verb second language,
that is, there should be exactly one constituent in front of the finite verb in declarative clauses
(leaving aside eliptical sentences). I developed analyses in the linearization"=based framework I
was working in, but for reasons that will be explained in this book, they were not satisfying. In
the end I changed my mind and completely revised my theories and computer implementations and
adopted a verb-movement analysis that is similar in spirit to the GB analysis. This analysis --
which was developed by \citet{Meurers99a}, based on work of \citet{KW91a,Kiss95a} -- is compatible
with the analysis of multiple frontings that is developed in this book. 

The present book is based on two articles that appeared in German in the journal \emph{Linguistische
  Berichte} in 2005 \citep{Mueller2005c,Mueller2005d}. Since these two papers belong to those of my papers
that are cited most often, I decided that it might be a good idea to make them availible to a wider
audience. The chapter~\ref{chap-german-sentence-structure} on German clause structure and the
Chapter~\ref{chapter-mult-front} on apparent multiple frontings and parts of
Chapter~\ref{chap-alternatives} on alternative analyses of the German clause are based on these
papers.  
The book ends with a chapter on empty elements, which is adapted from \citew{Mueller2004e}. This
chapter is meant to be a general discussion that shows what the cost is of alternative approaches
that try to avoid empty elements. Finally, there is an Appendix continaing a list of example
sentences that are used as a test suite for testing the computer-processable grammar that covers the
phenomena described in this book.



\section*{Acknowledgements}

I would like to thank the following people for their helpful comments and insights:
Bettina Braun,
Veronika Ehrich,
Gisbert Fanselow,
Peter Gallmann,
Rosemarie Lühr,
Detmar Meurers,
Susan Olsen,
Marga Reis, 
Christine Römer and 
Jan-Philipp Söhn 
as well as the anonymous reviewers from Formal Grammar, Konvens, CSLI Publications and \emph{Linguistische Berichte}.
I would also like to thank Anette Frank, Hans-Martin Gärtner, Tibor Kiss and Karel Oliva for their helpful discussion.

I have presented the analysis developed in this book at Formal Grammar 2002 in Trento, Konvens
2002 in Saarbrücken, at an invited talk for the SFB 441 at the University of Tübingen in 2002, at
the HPSG"=Workshop for Germanic languages 2003 at the University of Bremen, 
%
the Workshop Deutsche Syntax: Empirie und Theorie 2004 in Gothenburg,
%
at the Institute for Linguistics of the University Leipzig in 2004,
%
at the Formal Grammar conference 2004 in Nancy,
the 	Zentrum für Allgemeine Sprachwissenschaft, Typologie und
Universalienforschung (ZAS) in Berlin 2006,
%
at the Institute for German Language in Wuppertal in 2011,
%
at the colloquium
\emph{Der Satzanfang im Deutschen: syntaktische, semantisch-​pragmatische und
  informationsstrukturelle Integration vs. Desintegration} 2011 in Paris,
%
and at the workshop \emph{Satztypen und Konstruktionen im Deutschen: Satztypen:
  lexikalisch oder konfigurational?} at the Johannes Gutenberg-​Universität Mainz in
2013.
%
Many thanks to those present at these events for the subsequent discussion and I would especially like to thank the respective people/institutions for inviting me. 

The data discussed in this book come, in the most part, from my own collection of material I have
read over the years. Examples from the Mannheimer Morgen, Frankfurter Rundschau, St.\ Galler
Tagblatt, Tiroler Tageszeitung and the Züricher Tagesanzeiger were retrieved from the German Reference Corpus (\citealp{DeReKo}, \url{http://ids-mannheim.de/DeReKo}). Additional examples can be found at
\url{http://hpsg.fu-berlin.de/~stefan/Pub/mehr-vf-ds.html} and in the database of annotated  examples that was constructed from DeReKo data by Felix \citet{Bildhauer2011a} in the project A6 of the Collaborative Research Center/SFB 632 (accessible at \url{http://hpsg.fu-berlin.de/Resources/MVB/}).%

I would like to thank Andrew Murphy for translating \citew{Mueller2005c} and \citew{Mueller2005d},
which are the core of the Chapters~\ref{chapter-introduction}--\ref{chapter-mult-front} and for
proofreading these parts.


%~\bigskip

~\medskip

\noindent
Berlin, \today\hfill Stefan Müller



\mainmatter

\include{gs-introduction}
\include{gs-sentence-structure}
\include{gs-mult-fronting}
\include{gs-clause-types}
\include{gs-informationstructure}
\include{gs-alternatives}
\include{gs-empty-elements}
\include{gs-conclusion}

\appendix

\include{gs-tsdb-examples}

%\input{gt-index-cross}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Backmatter                       %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter

%\bibliography{bib-abbr,biblio,crossrefs}
\printbibliography[heading=references] 
\cleardoublepage

\small
   

\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{chapter}{Index} 
\addcontentsline{toc}{section}{Name index}
\ohead{Name index}
%with biblatex
%\printindex
%without it 
\printindex 
  
\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{section}{Language index}
\ohead{Language index} 
\printindex[lan] 
  
\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{section}{Subject index}
\ohead{Subject index}
\printindex[sbj]


\end{document} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%                  END                         %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%





%      <!-- Local IspellDict: en_US-w_accents -->
